Why do species get a thin slice of $\pi$? Revisiting Lewontin's Paradox of Variation

### 2021-05-27 manuscript/manuscript.tex
revised ms

 Neutral theory predicts that genetic diversity increases with population size,
 yet observed levels of diversity across metazoans vary only two orders of
 magnitude while population sizes vary over several. This unexpectedly narrow
 range of diversity is known as Lewontin's Paradox of Variation (1974). While
 some have suggested selection could constrain diversity, tests of this
 hypothesis seem to fall short. Here, I revisit Lewontin's Paradox and assess
 whether current models of linked selection are capable of reducing
 diversity to this extent. To quantify the discrepancy between pairwise
 diversity and census population sizes across species, I combine
 previously-published estimates of pairwise diversity from 172 metazoan taxa
 with estimates of census sizes. Using phylogenetic comparative methods, I show
 this relationship is significant accounting for phylogeny, but with high
 phylogenetic signal and evidence that some lineages experience shifts in the
 evolutionary rate of diversity deep in the past. Additionally, I find a
 negative relationship between recombination map length and census size,
 suggesting abundant species have less recombination and experience greater
 reductions in diversity due to linked selection. However, I show that even
 using strong selection parameter estimates, models of linked selection are
 unlikely to explain the observed relationship between diversity and census
 sizes across species.

@@

While chance demographic events like bottlenecks and recent expansions have
long-term impacts on diversity (since mutation-drift equilibrium is reached on
the order of size of the population), characteristics of the breeding structure
such as high variance ($V_w$) or skew in reproductive success  also suppress
diversity below the levels predicted by the census size
\parencite{Wright1938-tv}. For example, in many marine animals,
females are highly fecund, and dispersing larvae face extremely low
survivorship, leading to high variance in reproductive success
\parencite{Waples2018-kb,Waples2013-wi,Hedgecock2011-ku,Hauser2008-fd}. Such
``sweepstakes" reproductive systems can lead to remarkably small ratios of
effective population size to census population size (e.g. $\nicefrac{N_e}{N_c}$
can range from $10^{-6}$--$10^{-2}$), since $\nicefrac{N_e}{N} \approx
\nicefrac{1}{V_w}$
\parencite{Hedgecock1994-gs,Wright1938-tv,Nunney1993-ef,Nunney1996-wy}, and
require multiple-merger coalescent processes to describe their genealogies
\parencite{Eldon2006-ui}. Overall, these reproductive systems diminish the
diversity in many species, but seem unlikely to explain Lewontin's Paradox
broadly across metazoans.

@@

An important class of theoretic selection models pertaining to Lewontin's
Paradox are recurrent hitchhiking models that decouple diversity from the
census population size. These models predict diversity when strongly selected
beneficial mutations regularly enter and sweep through the population, trapping
lineages and forcing them to coalesce
\parencite{Kaplan1989-sc,Gillespie2000-mh}. In general, decoupling occurs under
these hitchhiking models when the rate of coalescence due to selection is much
greater than the rate of neutral coalescence \parencite[e.g.][equation
22]{Coop2012-cd}. Other linked selection models are proportional to population
size, and thus cannot decouple diversity, all else equal. For example, models
of background selection and polygenic fitness variation predict diversity is
proportional to population size, mediated by the total recombination map length
and the deleterious mutation rate or fitness variation
\parencite{Charlesworth1993-gb,Nicolaisen2012-vs,Nordborg1996-nq,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}.

@@

Past work looking at the relationship between $\pi$ and $N_c$ has been unable
to account for phylogenetic non-independence across taxa
\parencite{Felsenstein1985-an}. To address this shortcoming, I use phylogenetic
comparative methods (PCMs) with a synthetic time-calibrated phylogeny to
account for shared phylogenetic history. Moreover, it is disputed whether
considering phylogenetic non-independence is necessary in population genetics,
since coalescent times are much less than divergence times
\parencite{Whitney2010-ud,Lynch2011-qv}. Using PCMs, I address this by
estimating the degree of phylogenetic signal in the diversity census size
relationship, and investigating how these traits evolve along the phylogeny.

@@

A major impediment in quantifying Lewontin's Paradox has been estimating census
population sizes across many taxa, especially for extremely abundant,
cosmopolitan species that define the upper limit of ranges. Previous work has
surveyed the literature for census size estimates
\parencite{Nei1984-zi,Soule1976-he,Frankham1996-yb}, or used range or body
size or qualitative categories as proxies for census size
\parencite{Corbett-Detig2015-gt,Leffler2012-zj}. To quantify the relationship
between genomic estimates of diversity and census population sizes, I first
approximate census population sizes for 172 metazoan taxa (Figure
\ref{fig:figure-nc}). My approach predicts population densities from body sizes
using a previously-observed linear relationship that holds across metazoans
(Supplementary Figure \ref{suppfig:damuth};
\cite{Damuth1981-st,Damuth1987-sg}). Then, from geographic occurrence data, I
estimate range sizes.  Finally, I estimate population size as the product of
these predicted densities and range estimates (see Methods:
\Newnameref{sec:methods-popsize}). Note that the relationship between
population density and body size is driven by energy budgets, and thus reflects
macroecological equilibria \parencite{Damuth1987-sg}. Consequently, population
sizes are underestimated for taxa like humans and their domesticated species,
and overestimated for species with anthropogenically reduced densities or
patchy ranges. For example, the population size of \emph{Lynx lynx} is likely
around 50,000 \parencite{Iucn2020-ap} which is around two orders of magnitude
smaller than my estimate. Additionally, the range size estimates do not
consider whether an area has unsuitable habitat, and thus may be overestimates
for species with particular niches or patchy habitats. While these methods to
estimate census size are crude and approximate, they can be efficiently
calculated for numerous taxa and are sufficient to estimate the scale of
Lewontin's Paradox.

@@

  \caption{An illustration of Lewontin's Paradox of Variation. Pairwise
    diversity (data from \cite{Leffler2012-zj}, \cite{Corbett-Detig2015-gt},
    and \cite{Romiguier2014-bp}), which varies around three orders of
    magnitude, shows a weak relationship with approximate population size,
    which varies over 12 orders of magnitude.  The shaded curve shows the range
    of expected neutral diversity if $N_e$ were to equal $N_c$ under the
    four-alleles model, $\log_{10}(\pi) = \logt(\theta) - \logt{(1 +
    \nicefrac{4\theta}{3})}$ where $\theta = 4N_c \mu$, for two mutation rates,
    $\mu = 10^{-8}$ and $\mu = 10^{-9}$, and the light gray dashed line
    represents the maximum pairwise diversity under the four alleles model. The
    dark gray dashed line is the OLS regression fit, and the blue
    dashed line is the regression fit using a phylogenetic mixed-effects model.
    Points are colored by phylum.}

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size or range size, but these studies relied on heterozygosity measured from
allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi,Leffler2012-zj}. Here, I
confirm these findings using pairwise diversity estimates from genomic sequence
data and the estimated census sizes (Figure \ref{fig:figure-1}). The pairwise
diversity estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 0.09 percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[0.08, 0.12]$; see also the OLS fit per-phyla,
Supplementary Figure \ref{suppfig:figure-1-ave}).

@@

Using a phylogenetic mixed-effects model
\parencite{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in
Stan \parencite{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the
linear relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa without missing data and present in
the synthetic chronogram. As with the linear regression, this relationship was
positive and significant (95\% credible interval $0.04, 0.11$), though somewhat
attenuated compared to the OLS estimates (Figure \ref{fig:figure-2}B). Since
the population size estimates are based on range and body mass, they are
essentially a composite trait; fitting phylogenetic mixed-effects models
separately on body mass and range indicates these have significant negative and
positive effects, respectively (Supplementary Figure
\ref{suppfig:figure-div-range-bodymass}; see also Supplementary Figure
\ref{suppfig:div_leffler} for the relationship between diversity and the range
categories of \cite{Leffler2012-zj}).

With the phylogenetic mixed-effects model, I also estimated the variance of the
phylogenetic effect ($\sigma_p^2$) and the residual variance ($\sigma_r^2$),
which can be used to estimate a measure of the phylogenetic signal, $\lambda =
\nicefrac{\sigma_p^2}{(\sigma_p^2 + \sigma_r^2)}$
(\cite{Lynch1991-tp,De_Villemereuil2014-kt}; see \cite{Freckleton2002-ly} for a
comparison to Pagel's $\lambda$).  If the relationship between diversity and
population size was free of shared phylogenetic history, $\lambda = 0$ and all
the variance could be explained by evolution on the tips; this is analogous to
Lynch's conjecture that coalescent times should be free of phylogenetic signal
(\citeyear{Lynch2011-qv}). In the relationship between population size and
diversity, the posterior mean of $\lambda = 0.67$ (90\% credible interval
$[0.59, 0.75]$) indicates that the majority of the variance perhaps might be
due to shared phylogenetic history (Figure \ref{fig:figure-2}B).

@@

Additionally, I have explored the rate of trait change through time using
node-height tests \parencite{Freckleton2006-ww}. Node-height tests regress the
absolute values of the standardized contrasts between lineages against the
branching time (since present) of these lineages. Under Brownian Motion (BM),
standardized contrasts are estimates of the rate of character evolution
\parencite{Felsenstein1985-an}; if a trait evolves under constant rate BM, this
relationship should be flat. For both diversity and population size,
node-height tests indicate a significant increase in the rate of evolution
towards the present (robust regression p-values 0.028 and 0.00070 respectively;
Figure \ref{fig:figure-2}C). Considering the constituents of the population
size estimate, range and body mass, separately, the rate of evolution of range
but not body mass shows a significant increase (p-value $1.9 \times 10^{-7}$)
towards the present.

@@

I quantify the effect of linked selection on diversity as the ratio of observed
diversity ($\pi$) to the estimated diversity in the absence of linked selection
($\pi_0$), $R = \nicefrac{\pi}{\pi_0}$. Here, $\pi_0$ would reflect demographic
history and non-heritable variation in reproductive success. There are two
difficulties in evaluating whether linked selection could resolve Lewontin's
Paradox. The first difficulty is that $\pi_0$ is unobserved.  Previous work has
estimated $\pi_0$ using methods that exploit the spatial heterogeneity in
recombination and functional density across the genome to fit linked selection
models that incorporate both hitchhiking and background selection
\parencite{Elyashiv2016-vt,Corbett-Detig2015-gt}. The second difficulty is
understanding of how $R$ varies across taxa, since we lack estimates of
critical model parameters for most species. Still, I can address a key
question: if diversity levels were determined by census sizes ($\pi_0 = 4N_c
\mu$), are the combined effects of background selection and recurrent
hitchhiking sufficient to reduce diversity to observed levels? Furthermore,
does the relationship between census size and predicted diversity under linked
selection across species, $\pi_{BGS+HH} = R \pi_0$, match the observed
relationship in Figure \ref{fig:figure-1}?

@@

Previous work has found that the impact of linked selection increases with
$N_c$ (\cite{Corbett-Detig2015-gt}; see also Supplementary Figure
\ref{suppfig:corbett_detig}A), and it is often thought that this is driven by
higher rates of adaptive substitutions in larger populations, despite equivocal
evidence \parencite{Galtier2016-dq}. However, there is another mechanism by
which species with larger population sizes might experience a greater impact of
linked selection: recombinational map length, $L$, is known to correlate with
body mass \parencite{Burt1987-tq} and thus varies inversely with population
size. As this is a critical parameter that determines the genome-wide impact of
both hitchhiking and background selection, I examine the relationship between
recombination map length ($L$) and census population size ($N_c$) across taxa,
using available estimates of map lengths across species
\parencite{Stapley2017-fs,Corbett-Detig2015-gt}. I find a significant
non-linear relationship using phylogenetic mixed-effects models (Figure
\ref{fig:figure-3}A; see Methods: \Newnameref{sec:methods-pcm}). There is also
a correlation between map length and genome size (Supplementary Figure
\ref{supfig:genome_size_maplength}) and genome size and population size
(Supplementary Figure \ref{supfig:genome_size_popsize}). Overall, the negative
relationship between map length and census size indicates linked selection is
expected to be stronger in short map length, high-$N_c$ species.

Then, I predict the expected diversity ($\pi_{BGS+HH}$) under background
selection and hitchhiking, were $N_e= N_c$, and assuming all species had the
rate of sweeps and strength of BGS as \emph{D. melanogaster}. Since neutral
mutation rates $\mu$ are unknown and vary across species, I calculate the range
of predicted $\pi_{BGS+HH}$ estimates for $\mu = 10^{-8}$--$10^{-9}$ (using the
four-alleles model, \cite{Tajima1996-rb}), and compare this to the observed
relationship between $\pi$ and $N_c$ in Figured \ref{fig:figure-3}B.  Under
these parameters, linked selection begins to appreciably depress diversity
around $N_c \approx 10^9$, since $S \approx 10^{-8}$--$10^{-9}$ and linked
selection dominates drift when $S > \nicefrac{1}{2N}$.  Overall, this reveals
two problems for the hypothesis that linked selection could solve Lewontin's
Paradox. First, low to mid-$N_c$ species (census sizes between
$10^{6}$--$10^{14}$) have sufficiently long map lengths that their diversity
levels are only moderately reduced by linked selection, leading to a wide gap
between predicted and observed diversity levels. For this not to be the case,
the parameters that determine the strength of background selection and
recurrent hitchhiking would need to be \emph{higher} among these species than
in \emph{Drosophila melanogaster}. This would require that the rate of adaptive
mutations or the deleterious mutation rate be orders of magnitude higher for
species within this range than in \emph{Drosophila}, which is incompatible with
the rate of adaptive substitutions across species \parencite{Galtier2016-dq}
and mutation rates \parencite{Lynch2010-ki}. Furthermore, linked selection has
been quantified in humans, which fall in this census size range, and has been
found to be relatively weak
\parencite{McVicker2009-ax,Hernandez2011-gs,Hellmann2008-ic,Cai2009-by,Boyko2008-tj}.
Second, while hitchhiking and BGS can reduce predicted diversity levels for
high-$N_c$ species ($N_c > 10^{14}$) to observed levels, this would imply
available estimates of $\pi_0$ are underestimated by several orders of
magnitude in \emph{Drosophila} (Supplementary Figured
\ref{suppfig:corbett_detig}B). The high reductions in $\pi$ predicted here
(compared to those of \cite{Elyashiv2016-vt}) are a result of using $N_c$,
rather than $N_e = \nicefrac{\pi_0}{4 \mu}$ in the denominator of Equation
\eqref{eqn:div}, which leads to a very high rate of sweeps in the population.
I do not consider selective interference, though the saturation of adaptive
substitutions per Morgan would only act to limit the reduction in diversity
\parencite{Weissman2012-aa}, and thus these results are conservative.
Overall, while linked selection could decouple diversity from census size for
high-$N_c$ species, recurrent hitchhiking and background selection seem
unlikely to explain the observed patterns of diversity across species under our
understanding of the range of parameter estimates.

@@

Nearly fifty years after Lewontin's description of the Paradox of Variation,
how evolutionary, life history, and ecological processes interact to constrain
diversity across taxa to a narrow range remains a mystery. Since Wright
(\citeyear{Wright1931-fl,Wright1938-tv}), population geneticists have
appreciated that various demographic processes shrink effective population
sizes compared to census sizes, yet it has remained unclear whether these
neutral processes alone can explain Lewontin's Paradox and across-taxa
diversity patterns. Alternatively, selective processes that act more strongly
in larger populations could account for the observed narrow range of diversity.
A critical first step to discerning the processes that act to transform census
sizes to diversity levels across species is characterizing the observed
$\pi$--$N_c$ relationship.

Here, for 172 taxa, I estimated the relationship between genomic estimates of
pairwise diversity and approximate census population size. Previous surveys
have used allozyme-based estimates, fewer taxa, or qualitative measures of
population size. My estimates of census population sizes are quite approximate,
since they use body size to predict density. An improved estimate might
consider vagility (as \cite{Soule1976-he} did), though this is harder to do
systematically across many taxa. Future work might also use other ecological
information, such as total biomass and estimated numbers of species within
phyla, to improve census size estimates \parencite{Bar-On2018-kc, Mora2011-wm}.
Still, it seems more accurate estimates would be unlikely to change the
qualitative findings here, which resemble those of early surveys
\parencite{Nei1984-zi,Soule1976-he}.

@@

Of the full dataset of 172 taxa with diversity and population size estimates, a
synthetic calibrated phylogeny was created for 166 species that appear in
phylogenies in DateLife project \parencite{OMeara2020-ds,Sanchez-Reyes2019-io}.
This calibrated synthetic phylogeny was then subset for the analyses based on
what species had complete trait data. The diversity-population size
relationship assessed by a linear phylogenetic mixed-effects model implemented
in Stan \parencite{Stan_Development_Team2020-ea}, according to the methods
described in (\cite{De_Villemereuil2014-kt}, see
\texttt{stan/phylo\_mm\_regression.stan} in the GitHub repository). This same
Stan model was used to estimate the same relationship between arthropod,
chordate, and mollusc subsets of the data (Supplementary Figure
\ref{suppfig:div_popsize_post_phyla}).

>>>>

 Neutral theory predicts that genetic diversity increases with population size,
 yet observed levels of diversity across metazoans vary only two orders of
 magnitude while population sizes vary over several. This unexpectedly narrow
 range of diversity is known as Lewontin's Paradox of Variation (1974). While
 some have suggested selection constrains diversity, tests of this hypothesis
 seem to fall short. Here, I revisit Lewontin's Paradox to assess whether
 current models of linked selection are capable of reducing diversity to this
 extent. To quantify the discrepancy between pairwise diversity and census
 population sizes across species, I combine previously-published estimates of
 pairwise diversity from 172 metazoan taxa with estimates of census sizes.
 Using phylogenetic comparative methods, I show this relationship is
 significant accounting for phylogeny, but with high phylogenetic signal and
 evidence that some lineages experience shifts in the evolutionary rate of
 diversity deep in the past. Additionally, I find a negative relationship
 between recombination map length and census size, suggesting abundant species
 have less recombination and experience greater reductions in diversity due to
 linked selection. However, I show that even using strong selection parameter
 estimates, models of linked selection are unlikely to explain the observed
 relationship between diversity and census sizes across species.

@@

While chance demographic events like bottlenecks and recent expansions have
long-term impacts on diversity (since mutation-drift equilibrium is reached on
the order of size of the population), characteristics of the breeding structure
such as high variance ($V_w$) or skew in reproductive success continuously
suppress diversity below the levels predicted by the census size
\parencite{Wright1938-tv}. For example, in many marine animals, females are
highly fecund, and dispersing larvae face extremely low survivorship, leading
to high variance in reproductive success
\parencite{Waples2018-kb,Waples2013-wi,Hedgecock2011-ku,Hauser2008-fd}. Such
``sweepstakes" reproductive systems can lead to remarkably small ratios of
effective population size to census population size (e.g. $\nicefrac{N_e}{N_c}$
can range from $10^{-6}$--$10^{-2}$), since $\nicefrac{N_e}{N} \approx
\nicefrac{1}{V_w}$
\parencite{Hedgecock1994-gs,Wright1938-tv,Nunney1993-ef,Nunney1996-wy}, and
require multiple-merger coalescent processes to describe their genealogies
\parencite{Eldon2006-ui}. Overall, these reproductive systems diminish the
diversity in many species, but seem unlikely to explain Lewontin's Paradox
broadly across metazoans.

@@

An important class of theoretic selection models pertaining to Lewontin's
Paradox are recurrent hitchhiking models that decouple diversity from the
census population size. These models predict diversity when strongly selected
beneficial mutations regularly enter and sweep through the population, trapping
lineages and forcing them to coalesce
\parencite{Kaplan1989-sc,Gillespie2000-mh}. In general, decoupling occurs under
these hitchhiking models when the rate of coalescence due to selection is much
greater than the rate of neutral coalescence \parencite[e.g.][equation
22]{Coop2012-cd}. Under other linked selection models, the resulting effective
population size is proportional to population size, and thus these models
cannot decouple diversity, all else equal. For example, models of background
selection and polygenic fitness variation predict diversity is proportional to
population size, mediated by the total recombination map length and the
deleterious mutation rate or fitness variation
\parencite{Charlesworth1993-gb,Nicolaisen2012-vs,Nordborg1996-nq,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs}.

@@

Past work looking at the relationship between $\pi$ and $N_c$ has been unable
to fully account for phylogenetic non-independence across taxa
\parencite{Felsenstein1985-an}. To address this shortcoming, I use phylogenetic
comparative methods (PCMs) with a synthetic time-calibrated phylogeny to
account for shared phylogenetic history. Moreover, it is disputed whether
considering phylogenetic non-independence is necessary in population genetics,
since coalescent times are much less than divergence times
\parencite{Whitney2010-ud,Lynch2011-qv}. Using PCMs, I address this by
estimating the degree of phylogenetic signal in the diversity census size
relationship, and investigating how these traits evolve along the phylogeny.

@@

A major impediment in quantifying Lewontin's Paradox has been estimating census
population sizes across many taxa, especially for extremely abundant,
cosmopolitan species that define the upper limit of ranges. Previous work has
surveyed the literature for census size estimates
\parencite{Nei1984-zi,Soule1976-he,Frankham1996-yb}, or used range, body size,
or qualitative categories as proxies for census size
\parencite{Corbett-Detig2015-gt,Leffler2012-zj}. To quantify the relationship
between genomic estimates of diversity and census population sizes, I first
approximate census population sizes for 172 metazoan taxa (Figure
\ref{fig:figure-nc}). My approach predicts population densities from body sizes
using a previously-observed linear relationship that holds across metazoans
(Supplementary Figure \ref{suppfig:damuth};
\cite{Damuth1981-st,Damuth1987-sg}). Then, from geographic occurrence data, I
estimate range sizes.  Finally, I estimate population size as the product of
these predicted densities and range estimates (see Methods:
\Newnameref{sec:methods-popsize}). Note that the relationship between
population density and body size is driven by energy budgets, and thus reflects
macroecological equilibria \parencite{Damuth1987-sg}. Consequently, population
sizes are underestimated for taxa like humans and their domesticated species,
and overestimated for species with anthropogenically reduced densities or
fragmented ranges. For example, the population size of \emph{Lynx lynx} is
likely around 50,000 \parencite{Iucn2020-ap} which is around two orders of
magnitude smaller than my estimate. Additionally, the range size estimates do
not consider whether an area has unsuitable habitat, and thus may be
overestimated for species with particular niches or patchy habitats. While
these methods to estimate census size are crude and approximate, they can be
efficiently calculated for numerous taxa and are sufficient to estimate the
scale of Lewontin's Paradox (see Supplementary Information
\Newnameref{supp:popsize-valid} for more on validation based on biomass and
other approaches).

@@

  \caption{An illustration of Lewontin's Paradox of Variation. Pairwise
    diversity (data from \cite{Leffler2012-zj}, \cite{Corbett-Detig2015-gt},
    and \cite{Romiguier2014-bp}), which varies around three orders of
    magnitude, shows a weak relationship with approximate population size,
    which varies over 12 orders of magnitude.  The shaded curve shows the range
    of expected neutral diversity if $N_e$ were to equal $N_c$ under the
    four-alleles model, $\log_{10}(\pi) = \logt(\theta) - \logt{(1 +
    \nicefrac{4\theta}{3})}$ where $\theta = 4N_c \mu$, for two mutation rates,
    $\mu = 10^{-8}$ and $\mu = 10^{-9}$, and the light gray dashed line
    represents the maximum pairwise diversity under the four alleles model. The
    dark gray dashed line is the OLS regression fit, and the blue dashed line
    is the regression fit using a phylogenetic mixed-effects model.  Points are
    colored by phylum. The species \emph{Equus ferus przewalskii} ($N_c \approx
  10^3$ and $\pi = 3.6 \times 10^{-3}$) was an outlier and excluded from this
figure for visual clarity.}

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size or range size, but these studies relied on heterozygosity measured from
allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi,Leffler2012-zj}. Here, I
confirm these findings using pairwise diversity estimates from genomic sequence
data and the estimated census sizes (Figure \ref{fig:figure-1}). The pairwise
diversity estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 13\% percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[12\%, 14\%]$, adjusted $R^2 = 0.26$; see also the OLS fit
per-phyla, Supplementary Figure \ref{suppfig:figure-1-ave}).

@@

Using a phylogenetic mixed-effects model
\parencite{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in
Stan \parencite{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the
linear relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa without missing data and present in
the synthetic chronogram. As with the linear regression, this relationship was
positive and significant (95\% credible interval $0.03, 0.11$), though somewhat
attenuated compared to the OLS estimates (Figure \ref{fig:figure-2}B). Since
the population size estimates are based on range and body mass, they are
essentially a composite trait; fitting phylogenetic mixed-effects models
separately on body mass and range indicates these have significant negative and
positive effects, respectively (Supplementary Figure
\ref{suppfig:figure-div-range-bodymass}; see also Supplementary Figure
\ref{suppfig:div_leffler} for the relationship between diversity and the range
categories of \cite{Leffler2012-zj}).

With the phylogenetic mixed-effects model, I also estimated the variance of the
phylogenetic effect ($\sigma_p^2$) and the residual variance ($\sigma_r^2$),
which can be used to estimate a measure of the phylogenetic signal, $\lambda =
\nicefrac{\sigma_p^2}{(\sigma_p^2 + \sigma_r^2)}$
(\cite{Lynch1991-tp,De_Villemereuil2014-kt}; see \cite{Freckleton2002-ly} for a
comparison to Pagel's $\lambda$).  If the relationship between diversity and
population size was free of shared phylogenetic history, $\lambda = 0$ and all
the variance could be explained by evolution on the tips; this is analogous to
Lynch's conjecture that coalescent times should be free of phylogenetic signal
(\citeyear{Lynch2011-qv}). In the relationship between population size and
diversity, the posterior mean of $\lambda = 0.67$ (90\% credible interval
$[0.58, 0.75]$) indicates that the majority of the variance perhaps might be
due to shared phylogenetic history (Figure \ref{fig:figure-2}B).

@@

Additionally, I have explored the rate of trait change through time using
node-height tests \parencite{Freckleton2006-ww}. Node-height tests regress the
absolute values of the standardized contrasts between lineages against the
branching time (since present) of these lineages. Under Brownian Motion (BM),
standardized contrasts are estimates of the rate of character evolution
\parencite{Felsenstein1985-an}; if a trait evolves under constant rate BM, this
relationship should be flat. For both diversity and population size,
node-height tests indicate a significant increase in the rate of evolution
towards the present (robust regression p-values 0.023 and 0.00018 respectively;
Figure \ref{fig:figure-2}C). Considering the constituents of the population
size estimate, range and body mass, separately, the rate of evolution of range
but not body mass shows a significant increase (p-value $1.03 \times 10^{-7}$)
towards the present.

@@

I quantify the effect of linked selection on diversity as the ratio of observed
diversity ($\pi$) to the estimated diversity in the absence of linked selection
($\pi_0$), $R = \nicefrac{\pi}{\pi_0}$. Here, $\pi_0$ would reflect only
demographic history and non-heritable variation in reproductive success. There
are two difficulties in evaluating whether linked selection could resolve
Lewontin's Paradox. The first difficulty is that $\pi_0$ is unobserved.
Previous work has estimated $\pi_0$ using methods that exploit the spatial
heterogeneity in recombination and functional density across the genome to fit
linked selection models that incorporate both hitchhiking and background
selection \parencite{Elyashiv2016-vt,Corbett-Detig2015-gt}. The second
difficulty is understanding of how $R$ varies across taxa, since we lack
estimates of critical model parameters for most species. Still, I can address a
key question: if diversity levels were determined by census sizes ($\pi_0 =
4N_c \mu$), are the combined effects of background selection and recurrent
hitchhiking sufficient to reduce diversity to observed levels? Furthermore,
does the relationship between census size and predicted diversity under linked
selection across species, $\pi_{BGS+HH} = R \pi_0$, match the observed
relationship in Figure \ref{fig:figure-1}?

@@

Previous work has found that the impact of linked selection increases with
$N_c$ (\cite{Corbett-Detig2015-gt}; see also Supplementary Figure
\ref{suppfig:corbett_detig}A), and it is often thought that this is driven by
higher rates of adaptive substitutions in larger populations, despite equivocal
evidence \parencite{Galtier2016-dq}. However, there is another mechanism by
which species with larger population sizes might experience a greater impact of
linked selection: recombinational map length, $L$, is known to correlate with
body mass \parencite{Burt1987-tq} and thus varies inversely with population
size. As this is a critical parameter that determines the genome-wide impact of
both hitchhiking and background selection, I examine the relationship between
recombination map length ($L$) and census population size ($N_c$) across taxa,
using available estimates of map lengths across species
\parencite{Stapley2017-fs,Corbett-Detig2015-gt}. I find a significant
non-linear relationship using phylogenetic mixed-effects models (Figure
\ref{fig:figure-3}A; see Methods: \Newnameref{sec:methods-pcm}). There is also
a correlation between map length and genome size (Supplementary Figure
\ref{supfig:genome_size_maplength}) and genome size and population size
(Supplementary Figure \ref{supfig:genome_size_popsize}). These findings are
consistent with both the hypothesis that non-adaptive processes increase genome
size in small-$N_e$ species \parencite{Lynch2003-bn} which in turn could
increase map lengths, as well as the hypothesis that map lengths are adaptively
longer to more efficiently select against deleterious alleles
\parencite{Roze2021-ya}. Overall, the negative relationship between map length
and census size indicates linked selection is expected to be stronger in short
map length, high-$N_c$ species.

Then, I predict the expected diversity ($\pi_{BGS+HH}$) under background
selection and hitchhiking, were $N_e= N_c$, and assuming all species had the
rate of sweeps and strength of BGS as \emph{D. melanogaster}. Since neutral
mutation rates $\mu$ are unknown and vary across species, I calculate the range
of predicted $\pi_{BGS+HH}$ estimates for $\mu = 10^{-8}$--$10^{-9}$ (using the
four-alleles model, \cite{Tajima1996-rb}), and compare this to the observed
relationship between $\pi$ and $N_c$ in Figured \ref{fig:figure-3}B.  Under
these parameters, linked selection begins to appreciably depress diversity
around $N_c \approx 10^9$, since $S \approx 10^{-8}$--$10^{-9}$ and linked
selection dominates drift when $S > \nicefrac{1}{2N}$.  Overall, this reveals
two problems for the hypothesis that linked selection could solve Lewontin's
Paradox. First, low to mid-$N_c$ species (census sizes between
$10^{4}$--$10^{10}$) have sufficiently long map lengths that their diversity
levels are only moderately reduced by linked selection, leading to a wide gap
between predicted and observed diversity levels. For this not to be the case,
the parameters that determine the strength of background selection and
recurrent hitchhiking would need to be \emph{higher} among these species than
in \emph{Drosophila melanogaster}. This would require that the rate of adaptive
mutations or the deleterious mutation rate be orders of magnitude higher for
species within this range than in \emph{Drosophila}, which is incompatible with
the rate of adaptive substitutions across species \parencite{Galtier2016-dq}
and mutation rates \parencite{Lynch2010-ki}. Furthermore, linked selection has
been quantified in humans, which fall in this census size range, and has been
found to be relatively weak
\parencite{McVicker2009-ax,Hernandez2011-gs,Hellmann2008-ic,Cai2009-by,Boyko2008-tj}.
Second, while hitchhiking and BGS can reduce predicted diversity levels for
high-$N_c$ species ($N_c > 10^{10}$) to observed levels, this would imply
available estimates of $\pi_0$ are underestimated by several orders of
magnitude in \emph{Drosophila} (Supplementary Figured
\ref{suppfig:corbett_detig}B). The high reductions in $\pi$ predicted here
(compared to those of \cite{Elyashiv2016-vt}) are a result of using $N_c$,
rather than $N_e = \nicefrac{\pi_0}{4 \mu}$ in the denominator of Equation
\eqref{eqn:div}, which leads to a very high rate of sweeps in the population.
I do not consider selective interference, though the saturation of adaptive
substitutions per Morgan would only act to limit the reduction in diversity
\parencite{Weissman2012-aa}, and thus these results are conservative.
Overall, while linked selection could decouple diversity from census size for
high-$N_c$ species, recurrent hitchhiking and background selection seem
unlikely to explain the observed patterns of diversity across species under our
understanding of the range of parameter estimates.

@@

Nearly fifty years after Lewontin's description of the Paradox of Variation,
how evolutionary, life history, and ecological processes interact to constrain
diversity across taxa to a narrow range remains a mystery. I revisit Lewontin's
Paradox by first characterizing the relationship between genomic estimates of
pairwise diversity and approximate census population size across 172 metazoan
species. Previous surveys have used allozyme-based estimates, fewer taxa, or
qualitative measures of population size. My estimates of census population
sizes are quite approximate, since they use body size to predict density. An
improved estimate might consider vagility (as \cite{Soule1976-he} did), though
this is harder to do systematically across many taxa. Future work might also
use other ecological information, such as total biomass, or species
distribution modeling to improve census size estimates
\parencite{Bar-On2018-kc, Mora2011-wm}.  Still, it seems more accurate
estimates would be unlikely to change the qualitative findings here, which
resemble those of early surveys \parencite{Nei1984-zi,Soule1976-he}.

@@

Of the full dataset of 172 taxa with diversity and population size estimates, a
synthetic calibrated phylogeny was created for 166 species that appear in
phylogenies in DateLife project \parencite{OMeara2020-ds,Sanchez-Reyes2019-io}.
This calibrated synthetic phylogeny was then subset for the analyses based on
what species had complete trait data. The diversity-population size
relationship assessed by a linear phylogenetic mixed-effects model implemented
in Stan \parencite{Stan_Development_Team2020-ea}, according to the methods
described in (\cite{De_Villemereuil2014-kt}, see
\texttt{stan/phylo\_mm\_regression.stan} in the GitHub repository). This same
Stan model was used to estimate the same relationship between arthropod,
chordate, and mollusc subsets of the data, though a reduced model was used for
the chordate subset due to identifiability issues leading to poor MCMC
convergence (Supplementary Figure \ref{suppfig:div_popsize_post_phyla}).
